---
title: "Cohort Assembly"
author: "Tyler Tong"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
---

# Load packages 
```{R}
#| message: false
#| warning: false
library(tidyverse)
library(readr)
library(gt)
library(purrr)
library(janitor)
```

# Load datasets
```{r}
#| message: false
#| warning: false
DATA_PATH = "/Users/tt/AiReadi/AiReadiSummer2026"
mini_path <- paste0(
  DATA_PATH,
  "/Data/mini"
)
full_path <- paste0(
  DATA_PATH,
  "/Data/full"
)
```

```{r}
#| message: false
#| warning: false
participants <- read_tsv(
  file.path(mini_path, "participants.tsv"),
  show_col_types = FALSE)
cgm_manifest <- read_tsv(
  file.path(mini_path, "cgm.tsv"),
  show_col_types = FALSE
)
activity_manifest <- read_tsv(
  file.path(mini_path, "activity.tsv"),
  show_col_types = FALSE
)
fullparticipants <-  read_tsv(
  file.path(full_path, "participants.tsv"),
  show_col_types = FALSE)

rename_study_group <- function(x) {
  case_match(
    x,
    "healthy" ~ "healthy",
    "pre_diabetes_lifestyle_controlled" ~ "pre_diabetes",
    "oral_medication_and_or_non_insulin_injectable_medication_controlled" ~ "oral_meds",
    "insulin_dependent" ~ "insulin"
  )
}

participants <- participants |>
  mutate(study_group = rename_study_group(study_group))

fullparticipants <- fullparticipants |>
  mutate(study_group = rename_study_group(study_group))
```

# Cohort Table
```{R}
#| message: false
#| warning: false
# clean data
cgmdata <- cgm_manifest |> 
  select(person_id, average_glucose_level_mg_dl)
activitydata <- activity_manifest |> 
  select(person_id, average_daily_activity)
```

```{R}
#| message: false
#| warning: false
# join data by person_id
data <- participants |> 
  left_join(cgmdata, by = join_by(person_id))
data <- data |> 
  left_join(activitydata, by = join_by(person_id))
```

```{R}
#| message: false
#| warning: false
# display
head(data)
```


# Mini Dataset Analysis
```{R}
#| message: false
#| warning: false
studygroups <- participants |> 
  count(study_group)
gt(studygroups)
```


```{R}
#| message: false
#| warning: false
participants |>
  janitor::tabyl(clinical_site, study_group) |>
  janitor::adorn_totals(c("row", "col")) |>
  gt::gt()
```


```{R}
#| message: false
#| warning: false
countcgm <- participants |> 
  count(wearable_blood_glucose)
countactivity <- participants |> 
  count(wearable_activity_monitor)
countclinical <- participants |> 
  count(clinical_data)
countcgmwearable <- participants |> 
  count(wearable_blood_glucose, wearable_activity_monitor)
countall <- participants |> 
  count(wearable_blood_glucose, wearable_activity_monitor, clinical_data)

counts <- transmute(countcgm, cgm = map_dbl(n, 1))
counts <- counts |> 
  cross_join(transmute(countactivity, activity = map_dbl(n, 1)))
counts <- counts |> 
  cross_join(transmute(countclinical, clinical = map_dbl(n, 1)))
counts <- counts |> 
  cross_join(transmute(countcgmwearable, cgmwearable = map_dbl(n, 1)))
counts <- counts |> 
  cross_join(transmute(countcgmwearable, cgmactivityclinical = map_dbl(n, 1)))

gt(counts)
```

```{R}
#| message: false
#| warning: false

modality_counts <- participants |>
  select(
    cardiac_ecg, 
    clinical_data, 
    environment, 
    retinal_flio,
    retinal_oct, 
    retinal_octa, 
    retinal_photography,
    wearable_activity_monitor, 
    wearable_blood_glucose
  ) |>
  mutate(across(everything(), as.logical)) |>
  pivot_longer(
    cols = everything(), 
    names_to = "Modality", 
    values_to = "Is_Available"
  ) |>
  filter(Is_Available == TRUE) |>
  count(Modality, name = "Frequency") |>
  mutate(Modality = fct_reorder(Modality, Frequency))
```


```{R}
#| message: false
#| warning: false
ggplot(modality_counts, aes(x = Frequency, y = Modality, fill = Modality)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = Frequency), hjust = -0.2, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
```


# Mini vs Full
- The mini dataset has `r nrow(participants)` participants, while the full dataset has `r format(nrow(fullparticipants), big.mark = ",", scientific = FALSE)` participants.
- Both the full and mini have the same study groups and clinical sites.
- The full data set has some gaps in availability, but the mini has full coverage.
- any code written for the mini data set should run the same on the full data set.

# Analysis Cohorts

## CGM Cohort

```{R}
cgmcohort <- fullparticipants |> filter(wearable_blood_glucose)
summary(cgmcohort)
```

- All participants with CGM data are in this cohort.
- Contains `r format(nrow(cgmcohort), big.mark = ",", scientific = FALSE)` participants.
- Represents all `r n_distinct(cgmcohort$study_group)` study groups.
- Can be used to study how glucose changes in different people who have different conditions, as well as how glucose affects clinical data.
- `r sum(!cgmcohort$wearable_activity_monitor, na.rm = TRUE)` participants in this cohort do not have wearable data, so this cohort alone cannot be used to study how lifestyle affects glucose for those participants.

## CGM + Wearable Cohort

```{R}
cgmwearablecohort <- fullparticipants |> filter(wearable_blood_glucose, wearable_activity_monitor)
summary(cgmwearablecohort)
```

- Participants who have both CGM and activity data are in this cohort.
- It contains `r format(nrow(cgmwearablecohort), big.mark = ",", scientific = FALSE)` participants.
- It can be used to study how glucose responds to lifestyle changes, as well as how that may affect clinical data.
- It may not be used as a representative sample of the real population. Further 
investigation is need to assess whether this cohort have selection bias towards 
those who provided activity data.

# Mini Dataset Analysis
```{R}
#| message: false
#| warning: false
data <- participants |> count(study_group)
gt(data)
```
```{R}
#| message: false
#| warning: false
data <- participants |> count(clinical_site, study_group)
gt(data, groupname_col = "clinical_site", row_group_as_column = TRUE)
```
```{R}
#| message: false
#| warning: false

countcgm <- participants |> count(wearable_blood_glucose)
countwearable <- participants |> count(wearable_activity_monitor)
countclinical <- participants |> count(clinical_data)
countcgmwearable <- participants |> count(wearable_blood_glucose, wearable_activity_monitor)
countall <- participants |> count(wearable_blood_glucose, wearable_activity_monitor, clinical_data)

counts <- transmute(countcgm, cgm = map_dbl(n, 1))
counts <- counts |> cross_join(transmute(countwearable, wearable = map_dbl(n, 1)))
counts <- counts |> cross_join(transmute(countclinical, clinical = map_dbl(n, 1)))
counts <- counts |> cross_join(transmute(countcgmwearable, cgmwearable = map_dbl(n, 1)))
counts <- counts |> cross_join(transmute(countcgmwearable, cgmwearableclinical = map_dbl(n, 1)))

gt(counts)
```
```{R}
#| message: false
#| warning: false

modality_counts <- participants |>
  select(
    cardiac_ecg, 
    clinical_data, 
    environment, 
    retinal_flio,
    retinal_oct, 
    retinal_octa, 
    retinal_photography,
    wearable_activity_monitor, 
    wearable_blood_glucose
  ) |>
  mutate(across(everything(), as.logical)) |>
  pivot_longer(
    cols = everything(), 
    names_to = "Modality", 
    values_to = "Is_Available"
  ) |>
  filter(Is_Available == TRUE) |>
  count(Modality, name = "Frequency") |>
  mutate(Modality = fct_reorder(Modality, Frequency))
```
```{R}
#| message: false
#| warning: false
ggplot(modality_counts, aes(x = Frequency, y = Modality, fill = Modality)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = Frequency), hjust = -0.2, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))
```
# Mini vs Full
- Mini only has `r dim(participants)` participants, while the full dataset has 2k+.
- Both the full and mini have the same study groups and clinical sites.
- The full data set has some gaps in availability, but the mini has full coverage.
- any code written for the mini data set should run the same on the full data set.

# Analysis Cohorts

## CGM Cohort

```{R}
cgmcohort <- fullparticipants |> filter(wearable_blood_glucose)
tbl_summary(cgmwearablecohort) |> add_n() |> add_stat_label()
```

- All participants with CGM data are in this cohort.
- Contains 2245 participants.
- Represents all 4 study groups.
- Can be used to study how glucose changes in different people who have different conditions, as well as how glucose affects clinical data.
- 80 of the participants in this cohort do not have wearable data, so it can't be used to study how lifestyle affects glucose.

## CGM + Wearable Cohort

```{R}
cgmwearablecohort <- fullparticipants |> filter(wearable_blood_glucose, wearable_activity_monitor)
tbl_summary(cgmwearablecohort) |> add_n() |> add_stat_label()
```

- Participants who have CGM and wearable data are in this cohort.
- Contains 2165 participants.
- Can be used to study how glucose responds to lifestyle changes, as well as how that may affect clinical data.
- Cannot be used as a representative sample of the real population, as it may have selection bias towards those who provided wearable data.
