---
title: "Week 4: Sleep and Next-Day Glucose"
author: "Tyler Tong"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
execute:
  warning: false
  message: false
---

# Preparation

## Load packages and locate inputs

```{r}
rm(list = ls())

library(conflicted)
library(tidyverse)
library(here)
library(jsonlite)
library(janitor)
library(knitr)
library(scales)
library(broom)

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor = element_blank()
    )
)

dataset_version <- "v3-mini"

data_dir <- paste0("/Users/tt/Library/CloudStorage/Box-Box/ai-readi/", dataset_version, "/dataset")

derived_dir <- here("practice", "derived")
dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)

sleep_input_path <- file.path(
  derived_dir,
  "week3_person_night_sleep.csv"
)

activity_input_path <- file.path(
  derived_dir,
  "week3_daily_activity.csv"
)

cgm_input_path <- file.path(
  derived_dir,
  "week2_participant_day_cgm.csv"
)

if (!file.exists(sleep_input_path) || !file.exists(activity_input_path)) {
  stop(
    "Run 03_sleep_activity_processing.qmd first; its two processed ",
    "tables are required for Week 4."
  )
}

resolve_data_path <- function(relative_path) {
  file.path(data_dir, str_remove(relative_path, "^/"))
}
```

## Define the analysis rules

```{r}
study_group_labels <- c(
  healthy = "Healthy",
  pre_diabetes_lifestyle_controlled = "Pre-DM / lifestyle",
  oral_medication_and_or_non_insulin_injectable_medication_controlled =
    "Oral/non-insulin medication",
  insulin_dependent = "Insulin-dependent"
)
```

# Read Week 3 tables and participant data

```{r}
person_night_sleep <- read_csv(
  sleep_input_path,
  show_col_types = FALSE
) |>
  mutate(
    person_id = as.character(person_id),
    sleep_night = as.Date(sleep_night),
    bedtime_utc = ymd_hms(bedtime_utc, tz = "UTC"),
    wake_time_utc = ymd_hms(wake_time_utc, tz = "UTC"),
    study_group_label = factor(
      study_group_label,
      levels = unname(study_group_labels)
    ), 
    date = date(wake_time_local)
  )

daily_activity <- read_csv(
  activity_input_path,
  show_col_types = FALSE
) |>
  mutate(
    person_id = as.character(person_id),
    date = as.Date(date),
    study_group_label = factor(
      study_group_label,
      levels = unname(study_group_labels)
    )
  )

person_day_cgm <- read_csv(
  cgm_input_path,
  show_col_types = FALSE
) |>
  mutate(
    person_id = as.character(person_id),
    date_local = as.Date(date_local)
  )

```

# Prepare and Join Inputs

## Prepare Inputs

```{r}
person_day_cgm <- person_day_cgm |> 
  rename(date = date_local) |>
  group_by(person_id, date) |>
  mutate(
    glucose_cv = (sd_glucose / mean_glucose) * 100
  ) |>
  ungroup()
person_night_sleep <- person_night_sleep |> mutate(
  bedtime_local = ymd_hms(bedtime_local), 
  sleep_onset_local = ymd_hms(sleep_onset_local), 
  final_sleep_end_local = ymd_hms(final_sleep_end_local),
  wake_time_local = ymd_hms(wake_time_local)
  )
daily_activity <- daily_activity |>
  select(!study_group) |> 
  select(!clinical_site) |>
  select(!study_group_label)
glimpse(daily_activity)
```

## Find unmatched participant-dates

```{r}
dfs <- list(person_day_cgm = person_day_cgm, person_night_sleep = person_night_sleep, daily_activity = daily_activity)
nms <- names(dfs)

key_long <- dfs |>
  map(\(d) d |>
        mutate(date = as.Date(date)) |>
        dplyr::filter(!is.na(person_id), !is.na(date)) |>
        distinct(person_id, date)) |>
  list_rbind(names_to = "source")

incomplete <- key_long |>
  summarise(sources = list(unique(source)), .by = c(person_id, "date")) |>
  dplyr::filter(lengths(sources) < length(nms)) |>
  mutate(
    present_in   = map_chr(sources, str_flatten, collapse = ", "),
    missing_from = map_chr(sources, \(s) str_flatten(setdiff(nms, s), ", "))
  ) |>
  select(person_id, date, present_in, missing_from) |>
  arrange(person_id, date)

glimpse(incomplete)

bad_keys <- incomplete |>
  select(person_id, date)

dfs <- dfs |>
  map(\(d) anti_join(d, bad_keys,
                     by = join_by(person_id, date == date)))
```

## Join `inputs` on `person_id` and `date`

```{r}
data <- dfs$person_day_cgm |>
  left_join(dfs$person_night_sleep, by = join_by(
    person_id == person_id,
    date == date
  )) |>
  left_join(dfs$daily_activity, by = join_by(
    person_id == person_id, 
    date == date
  ))
glimpse(data)
```

# Build the Participant-level Analysis Table

## Build the table from `data` using `summarize()`

```{r}
participant_table <- data |>
  ungroup() |>
  group_by(person_id) |>
  summarize(
    num_days                     = n(),
    typical_sleep_dur_min        = median(total_sleep_time_minutes),
    typical_daily_steps          = median(total_steps),
    typical_daily_glucose_cv     = median(glucose_cv),
    typical_daily_time_in_range  = median(pct_70_180) * 14.40,
    .groups = "drop"
  )
```

We use `median()` here because the data is likely to be skewed and asymmetric, making it less suitable for `mean()`.

## Preview the analysis table

```{r}
kable(
  head(participant_table), 
  col.names = c("ID", 
                "Days", 
                "Typical Sleep Duration", 
                "Typical Daily Steps", 
                "Typical Daily Glucose CV", 
                "Typical Daily Time In Range"), 
  align = "cccccc", 
  digits = 2, 
  caption = "First Six Rows of the Participant Analysis Table"
  )
```

# Create Participant Level Plots

## Seep Duration vs. Glucose CV

```{r}
#| message: false
#| warning: false
participant_table |>
ggplot(
  aes(
    x = typical_sleep_dur_min,
    y = typical_daily_glucose_cv
  )
) + 
  geom_point() + 
  geom_smooth() + 
  labs(x = "Typical Sleep Duration (minutes)", y = "Typical Daily Glucose CV")

correlation_data_sleep <- tidy(cor.test(
  participant_table$typical_sleep_dur_min, 
  participant_table$typical_daily_glucose_cv, 
  method = "spearman"
  )) |>
  select(estimate, p.value) |>
  mutate(num_participants = nrow(participant_table))

kable(
  correlation_data_sleep, 
  col.names = c("Spearman Estimate", "P Value", "Participant Count"),
  align = "ccc",
  digits = 3
  )
```

As seen in the `geom_smooth()` line, the relationship between sleep duration and glucose CV is very small. This is also reflected by the Spearman estimate, `{r} correlation_data_sleep$estimate`, which is near zero, showing that any linear correlation, if extant, is very weak. In addition, the p value is `{r} correlation_data_sleep$p.value`, which is near 1, meaning that there is no strong evidence that any correlation exists. The sample size is also relatively large, `{r} correlation_data_sleep$num_participants`, so the results should not be because of a small sample size.

## Daily Steps vs. Glucose CV

```{r}
#| message: false
#| warning: false
participant_table |>
ggplot(
  aes(
    x = typical_daily_steps,
    y = typical_daily_glucose_cv
  )
) + 
  geom_point() + 
  geom_smooth() + 
  labs(x = "Typical Daily Steps", y = "Typical Daily Glucose CV")

correlation_data_steps <- tidy(cor.test(
  participant_table$typical_daily_steps, 
  participant_table$typical_daily_glucose_cv, 
  method = "spearman"
  )) |>
  select(estimate, p.value) |>
  mutate(num_participants = nrow(participant_table))

kable(
  correlation_data_steps, 
  col.names = c("Spearman Estimate", "P Value", "Participant Count"),
  align = "ccc",
  digits = 3
  )
```

Here, The `geom_smooth()` line is mostly flat, but it shows a bump where the data points are densest. This indicates a very small correlation, and the Spearman estimate confirms this, with `{r} correlation_data_steps$estimate` being very close to zero. The p-value, `{r} correlation_data_steps$p.value`, is very close to being significant, but falls short. This indicates that the relationship has a possibility of existing, but it cannot be determined that it is not random luck. Again, the participant count of `{r} correlation_data_steps$num_participants` should be enough to confirm any relationships within the data.

# Interpretation

## Which association is stronger in absolute magnitude?

The association between daily steps and glucose CV is stronger than that between sleep duration and glucose CV(`{r} abs(correlation_data_steps$estimate)` > `{r} abs(correlation_data_sleep$estimate)`).

## What does the direction of each correlation mean?

The association of daily steps and glucose CV is negative(`{r} correlation_data_steps$estimate` < 0), meaning that as daily steps increase, glucose CV should decrease. On the other hand, the association between sleep duration and glucose CV is positive(`{r} correlation_data_sleep$estimate` > 0), meaning that as sleep duration increases, glucose CV should increase.