---
title: "Week 4: Sleep and Next-Day Glucose"
author: "Tyler Tong"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
execute:
  warning: false
  message: false
---

# Preparation

## Load packages and locate inputs

```{r}
rm(list = ls())

library(tidyverse)
library(here)
library(jsonlite)
library(janitor)
library(knitr)
library(scales)

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor = element_blank()
    )
)

dataset_version <- "v3-mini"

data_dir <- paste0("/Users/tt/Library/CloudStorage/Box-Box/ai-readi/", dataset_version, "/dataset")

derived_dir <- here("practice", "derived")
dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)

sleep_input_path <- file.path(
  derived_dir,
  "week3_person_night_sleep.csv"
)

activity_input_path <- file.path(
  derived_dir,
  "week3_daily_activity.csv"
)

if (!file.exists(sleep_input_path) || !file.exists(activity_input_path)) {
  stop(
    "Run 03_sleep_activity_processing.qmd first; its two processed ",
    "tables are required for Week 4."
  )
}

resolve_data_path <- function(relative_path) {
  file.path(data_dir, str_remove(relative_path, "^/"))
}
```

## Define the analysis rules

```{r}
next_day_window_hours <- 18
expected_cgm_readings_per_hour <- 12
minimum_cgm_coverage <- 0.70
minimum_nights_per_participant <- 4
minimum_sleep_minutes_for_analysis <- 2 * 60

expected_readings_per_window <-
  next_day_window_hours * expected_cgm_readings_per_hour

minimum_readings_per_window <- ceiling(
  minimum_cgm_coverage * expected_readings_per_window
)

study_group_labels <- c(
  healthy = "Healthy",
  pre_diabetes_lifestyle_controlled = "Pre-DM / lifestyle",
  oral_medication_and_or_non_insulin_injectable_medication_controlled =
    "Oral/non-insulin medication",
  insulin_dependent = "Insulin-dependent"
)
```

# Read Week 3 tables and participant data

```{r}
person_night_sleep <- read_csv(
  sleep_input_path,
  show_col_types = FALSE
) |>
  mutate(
    person_id = as.character(person_id),
    sleep_night = as.Date(sleep_night),
    bedtime_utc = ymd_hms(bedtime_utc, tz = "UTC"),
    wake_time_utc = ymd_hms(wake_time_utc, tz = "UTC"),
    study_group_label = factor(
      study_group_label,
      levels = unname(study_group_labels)
    )
  )

daily_activity <- read_csv(
  activity_input_path,
  show_col_types = FALSE
) |>
  mutate(
    person_id = as.character(person_id),
    date = as.Date(date),
    study_group_label = factor(
      study_group_label,
      levels = unname(study_group_labels)
    )
  )

participants <- read_tsv(
  file.path(data_dir, "participants.tsv"),
  show_col_types = FALSE
) |>
  clean_names() |>
  mutate(person_id = as.character(person_id))

cgm_manifest <- read_tsv(
  file.path(data_dir, "wearable_blood_glucose", "manifest.tsv"),
  show_col_types = FALSE
) |>
  clean_names() |>
  mutate(person_id = as.character(person_id))
```

# Read and clean CGM

## Read all CGM participants

```{r}
read_cgm_file <- function(person_id, relative_path) {
  cgm_json <- fromJSON(
    resolve_data_path(relative_path),
    flatten = TRUE
  )

  cgm_body <- cgm_json$body$cgm

  tibble(
    person_id = as.character(person_id),
    time_utc = ymd_hms(
      cgm_body$effective_time_frame.time_interval.start_date_time,
      tz = "UTC",
      quiet = TRUE
    ),
    event_type = as.character(cgm_body$event_type),
    glucose_original = as.character(cgm_body$blood_glucose.value),
    glucose_mg_dl = suppressWarnings(
      as.numeric(glucose_original)
    ),
    glucose_unit = as.character(cgm_body$blood_glucose.unit)
  ) |>
    filter(event_type == "EGV")
}

cgm_raw_egv <- map2_dfr(
  cgm_manifest$person_id,
  cgm_manifest$glucose_filepath,
  read_cgm_file
)
```

## Apply the Week 2 cleaning rules needed here

```{r}
cgm_clean <- cgm_raw_egv |>
  filter(
    !is.na(time_utc),
    !is.na(glucose_mg_dl),
    glucose_unit == "mg/dL",
    between(glucose_mg_dl, 40, 400)
  ) |>
  group_by(person_id, time_utc) |>
  filter(n_distinct(glucose_mg_dl) == 1) |>
  slice_head(n = 1) |>
  ungroup() |>
  arrange(person_id, time_utc)

tibble(
  check = c(
    "CGM participants",
    "Raw EGV records",
    "Usable unique CGM readings",
    "Removed or collapsed records"
  ),
  result = c(
    n_distinct(cgm_clean$person_id),
    nrow(cgm_raw_egv),
    nrow(cgm_clean),
    nrow(cgm_raw_egv) - nrow(cgm_clean)
  )
) |>
  kable(
    col.names = c("CGM processing check", "Result"),
    caption = "CGM records available for post-wake windows"
  )
```

nonnumeric Dexcom `"Low"`/`"High"` strings, missing timestamps, non-mg/dL units, values outside the exercise bounds of 40--400 mg/dL, and conflicting duplicate timestamps are not used.

# Build the linked person-night table

## Summarize CGM after each wake time

```{r}
summarise_next_day_cgm <- function(person_id, wake_time_utc) {
  window_end_utc <- wake_time_utc + hours(next_day_window_hours)

  window_readings <- cgm_clean |>
    filter(
      .data$person_id == person_id,
      time_utc >= wake_time_utc,
      time_utc < window_end_utc
    )

  number_of_readings <- nrow(window_readings)
  coverage_percent <-
    100 * number_of_readings / expected_readings_per_window
  valid_window <-
    number_of_readings >= minimum_readings_per_window

  if (!valid_window) {
    return(
      tibble(
        next_day_window_end_utc = window_end_utc,
        next_day_cgm_readings = number_of_readings,
        next_day_cgm_coverage_percent = coverage_percent,
        valid_next_day_cgm = FALSE,
        next_day_mean_glucose = NA_real_,
        next_day_peak_glucose = NA_real_,
        next_day_glucose_sd = NA_real_,
        next_day_glucose_cv_percent = NA_real_,
        next_day_time_in_range_percent = NA_real_,
        next_day_time_above_180_percent = NA_real_,
        next_day_time_below_70_percent = NA_real_
      )
    )
  }

  mean_glucose <- mean(window_readings$glucose_mg_dl)
  glucose_sd <- sd(window_readings$glucose_mg_dl)

  tibble(
    next_day_window_end_utc = window_end_utc,
    next_day_cgm_readings = number_of_readings,
    next_day_cgm_coverage_percent = coverage_percent,
    valid_next_day_cgm = TRUE,
    next_day_mean_glucose = mean_glucose,
    next_day_peak_glucose = max(window_readings$glucose_mg_dl),
    next_day_glucose_sd = glucose_sd,
    next_day_glucose_cv_percent =
      100 * glucose_sd / mean_glucose,
    next_day_time_in_range_percent = 100 * mean(
      between(window_readings$glucose_mg_dl, 70, 180)
    ),
    next_day_time_above_180_percent = 100 * mean(
      window_readings$glucose_mg_dl > 180
    ),
    next_day_time_below_70_percent = 100 * mean(
      window_readings$glucose_mg_dl < 70
    )
  )
}

sleep_cgm_candidates <- person_night_sleep |>
  filter(person_id %in% cgm_manifest$person_id) |>
  mutate(
    sleep_data_quality_ok =
      total_sleep_time_minutes >= minimum_sleep_minutes_for_analysis
  )

next_day_cgm_features <- pmap_dfr(
  sleep_cgm_candidates |>
    select(person_id, wake_time_utc),
  summarise_next_day_cgm
)

linked_person_night <- bind_cols(
  sleep_cgm_candidates,
  next_day_cgm_features
) |>
  left_join(
    daily_activity |>
      select(
        person_id,
        activity_date = date,
        total_steps,
        active_calories,
        average_heart_rate_bpm,
        step_observation_hours,
        valid_activity_day
      ),
    by = c(
      "person_id",
      "sleep_night" = "activity_date"
    )
  ) |>
  mutate(
    activity_available = !is.na(total_steps),
    valid_activity_day = coalesce(valid_activity_day, FALSE)
  ) |>
  arrange(person_id, sleep_night)

linked_output_path <- file.path(
  derived_dir,
  "week4_sleep_next_day_linked.csv"
)

write_csv(linked_person_night, linked_output_path, na = "")
```

## Linkage checks

```{r}
linked_key_duplicates <- linked_person_night |>
  count(person_id, sleep_night) |>
  filter(n > 1) |>
  nrow()

tibble(
  check = c(
    "Sleep nights among CGM participants",
    "Nights with analysis-quality sleep",
    "Nights with valid next-day CGM",
    "Nights with complete same-day activity",
    "Participants with any valid linked night",
    "Duplicate person-night keys"
  ),
  result = c(
    nrow(linked_person_night),
    sum(linked_person_night$sleep_data_quality_ok),
    sum(linked_person_night$valid_next_day_cgm),
    sum(linked_person_night$valid_activity_day),
    linked_person_night |>
      filter(
        sleep_data_quality_ok,
        valid_next_day_cgm
      ) |>
      summarise(n = n_distinct(person_id)) |>
      pull(n),
    linked_key_duplicates
  )
) |>
  kable(
    col.names = c("Linkage check", "Result"),
    caption = "Person-night linkage quality control"
  )
```

## Preview the analysis table

```{r}
linked_person_night |>
  filter(valid_next_day_cgm) |>
  select(
    person_id,
    sleep_night,
    total_sleep_time_minutes,
    sleep_efficiency_percent,
    awake_after_sleep_onset_minutes,
    next_day_cgm_readings,
    next_day_cgm_coverage_percent,
    next_day_mean_glucose,
    next_day_peak_glucose,
    next_day_time_in_range_percent,
    total_steps,
    active_calories,
    average_heart_rate_bpm,
    valid_activity_day,
    study_group_label
  ) |>
  slice_head(n = 12) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 1)
    )
  ) |>
  kable(
    col.names = c(
      "Participant",
      "Wake date",
      "TST (min)",
      "Efficiency (%)",
      "WASO (min)",
      "CGM readings",
      "CGM coverage (%)",
      "Next-day mean",
      "Next-day peak",
      "Next-day TIR (%)",
      "Steps",
      "Active calories",
      "Average HR",
      "Complete activity",
      "Study group"
    ),
    caption = "First 12 valid linked person-nights"
  )
```

The complete linkage table retains nights with insufficient CGM or incomplete activity and marks their status rather than deleting them. It is saved as `practice/derived/week4_sleep_next_day_linked.csv`.

# Define worse and better sleep within participants

```{r}
analysis_nights <- linked_person_night |>
  filter(
    sleep_data_quality_ok,
    valid_next_day_cgm
  ) |>
  group_by(person_id) |>
  filter(n() >= minimum_nights_per_participant) |>
  mutate(
    person_median_sleep_minutes = median(
      total_sleep_time_minutes,
      na.rm = TRUE
    ),
    sleep_group = if_else(
      total_sleep_time_minutes < person_median_sleep_minutes,
      "worse",
      "better"
    )
  ) |>
  filter(n_distinct(sleep_group) == 2) |>
  ungroup() |>
  mutate(
    sleep_group = factor(
      sleep_group,
      levels = c("better", "worse")
    ),
    study_group_label = factor(
      study_group_label,
      levels = unname(study_group_labels)
    )
  )

paired_sleep_summary <- analysis_nights |>
  group_by(
    person_id,
    study_group,
    study_group_label,
    sleep_group
  ) |>
  summarise(
    number_of_nights = n(),
    mean_sleep_hours =
      mean(total_sleep_time_minutes) / 60,
    next_day_mean_glucose =
      mean(next_day_mean_glucose),
    next_day_peak_glucose =
      mean(next_day_peak_glucose),
    next_day_time_in_range_percent =
      mean(next_day_time_in_range_percent),
    next_day_time_above_180_percent =
      mean(next_day_time_above_180_percent),
    .groups = "drop"
  )

paired_differences <- paired_sleep_summary |>
  pivot_wider(
    names_from = sleep_group,
    values_from = c(
      number_of_nights,
      mean_sleep_hours,
      next_day_mean_glucose,
      next_day_peak_glucose,
      next_day_time_in_range_percent,
      next_day_time_above_180_percent
    )
  ) |>
  filter(
    !is.na(next_day_mean_glucose_better),
    !is.na(next_day_mean_glucose_worse)
  ) |>
  mutate(
    sleep_duration_gap_hours =
      mean_sleep_hours_better - mean_sleep_hours_worse,
    mean_glucose_difference =
      next_day_mean_glucose_worse -
        next_day_mean_glucose_better,
    peak_glucose_difference =
      next_day_peak_glucose_worse -
        next_day_peak_glucose_better,
    time_in_range_difference =
      next_day_time_in_range_percent_worse -
        next_day_time_in_range_percent_better,
    time_above_180_difference =
      next_day_time_above_180_percent_worse -
        next_day_time_above_180_percent_better
  )

paired_output_path <- file.path(
  derived_dir,
  "week4_paired_participant_differences.csv"
)

write_csv(paired_differences, paired_output_path, na = "")
```

The primary difference is:

> Mean next-day glucose after worse-sleep nights minus mean next-day glucose after better-sleep nights.

A positive difference means higher next-day mean glucose after shorter sleep for that participant. A negative difference means lower next-day mean glucose after shorter sleep.

```{r}
paired_differences |>
  select(
    person_id,
    study_group_label,
    number_of_nights_better,
    number_of_nights_worse,
    mean_sleep_hours_better,
    mean_sleep_hours_worse,
    next_day_mean_glucose_better,
    next_day_mean_glucose_worse,
    mean_glucose_difference
  ) |>
  slice_head(n = 15) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 1)
    )
  ) |>
  kable(
    col.names = c(
      "Participant",
      "Study group",
      "Better nights",
      "Worse nights",
      "Better sleep (h)",
      "Worse sleep (h)",
      "Glucose after better",
      "Glucose after worse",
      "Difference (worse - better)"
    ),
    caption = "First 15 within-participant paired summaries"
  )
```

The complete paired table contains `r nrow(paired_differences)` participants and is saved as `practice/derived/week4_paired_participant_differences.csv`.

# Required plots

## Slopegraph: better versus worse sleep

```{r}
#| fig-width: 9
#| fig-height: 6

paired_sleep_summary |>
  filter(person_id %in% paired_differences$person_id) |>
  mutate(
    sleep_group = factor(
      sleep_group,
      levels = c("better", "worse"),
      labels = c("Better sleep", "Worse sleep")
    )
  ) |>
  ggplot(
    aes(
      x = sleep_group,
      y = next_day_mean_glucose,
      group = person_id
    )
  ) +
  geom_line(
    color = "#7A7F87",
    alpha = 0.65,
    linewidth = 0.7
  ) +
  geom_point(
    aes(color = study_group_label),
    size = 2.7
  ) +
  scale_color_manual(
    values = c(
      "Healthy" = "#2C6E9B",
      "Pre-DM / lifestyle" = "#C47B22",
      "Oral/non-insulin medication" = "#8A3F69",
      "Insulin-dependent" = "#6B7D3A"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Next-day mean glucose after better- and worse-sleep nights",
    subtitle = "Each line is one participant; sleep groups are defined by that participant's own median total sleep time.",
    x = NULL,
    y = "Mean glucose in the 18-hour post-wake window (mg/dL)",
    color = "Study group"
  ) +
  theme(legend.position = "bottom")
```

## Boxplot of within-person glucose differences

```{r}
#| fig-width: 7
#| fig-height: 5

set.seed(20260728)

paired_differences |>
  ggplot(
    aes(
      x = "All paired participants",
      y = mean_glucose_difference
    )
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "#555555"
  ) +
  geom_boxplot(
    width = 0.35,
    fill = "#DDEAF7",
    color = "#2C6E9B",
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    color = "#2C6E9B",
    size = 2.3,
    alpha = 0.8
  ) +
  labs(
    title = "Within-person difference in next-day mean glucose",
    subtitle = str_wrap(
      "Difference = worse-sleep minus better-sleep; positive values indicate higher glucose after shorter sleep.",
      width = 80
    ),
    x = NULL,
    y = "Mean glucose difference (mg/dL)"
  )
```

## Differences stratified by study group

```{r}
#| fig-width: 10
#| fig-height: 6

set.seed(20260728)

paired_differences |>
  ggplot(
    aes(
      x = study_group_label,
      y = mean_glucose_difference
    )
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "#555555"
  ) +
  geom_boxplot(
    width = 0.5,
    fill = "#E9EEF3",
    color = "#454B52",
    outlier.shape = NA
  ) +
  geom_jitter(
    aes(color = study_group_label),
    width = 0.12,
    height = 0,
    size = 2.5,
    alpha = 0.85
  ) +
  scale_color_manual(
    values = c(
      "Healthy" = "#2C6E9B",
      "Pre-DM / lifestyle" = "#C47B22",
      "Oral/non-insulin medication" = "#8A3F69",
      "Insulin-dependent" = "#6B7D3A"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Within-person glucose differences by AI-READI study group",
    subtitle = "Points are participants, not nights; small group counts make this descriptive only.",
    x = NULL,
    y = "Mean glucose difference, worse - better sleep (mg/dL)"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 18, hjust = 1)
  )
```

## Example participant: sleep nights and next-day glucose

```{r}
example_participant <- analysis_nights |>
  count(person_id, name = "number_of_eligible_nights") |>
  arrange(desc(number_of_eligible_nights), person_id) |>
  slice_head(n = 1) |>
  pull(person_id)

example_nights <- analysis_nights |>
  filter(person_id == example_participant) |>
  arrange(sleep_night) |>
  mutate(
    monitoring_episode = cumsum(
      is.na(lag(sleep_night)) |
        as.numeric(sleep_night - lag(sleep_night)) > 14
    )
  )

example_plot_data <- bind_rows(
  example_nights |>
    transmute(
      sleep_night,
      sleep_group,
      monitoring_episode,
      metric = "Previous-night total sleep time (hours)",
      value = total_sleep_time_minutes / 60
    ),
  example_nights |>
    transmute(
      sleep_night,
      sleep_group,
      monitoring_episode,
      metric = "Next-day mean glucose (mg/dL)",
      value = next_day_mean_glucose
    )
) |>
  mutate(
    metric = factor(
      metric,
      levels = c(
        "Previous-night total sleep time (hours)",
        "Next-day mean glucose (mg/dL)"
      )
    )
  )

example_sleep_reference <- tibble(
  metric = factor(
    "Previous-night total sleep time (hours)",
    levels = levels(example_plot_data$metric)
  ),
  reference = unique(
    example_nights$person_median_sleep_minutes
  ) / 60
)
```

```{r}
#| fig-width: 10
#| fig-height: 7

example_plot_data |>
  ggplot(
    aes(
      x = sleep_night,
      y = value
    )
  ) +
  geom_hline(
    data = example_sleep_reference,
    aes(yintercept = reference),
    linetype = "dashed",
    color = "#555555",
    inherit.aes = FALSE
  ) +
  geom_line(
    aes(group = interaction(metric, monitoring_episode)),
    color = "#8A9097",
    linewidth = 0.6
  ) +
  geom_point(
    aes(color = sleep_group),
    size = 3
  ) +
  facet_wrap(
    ~ metric,
    ncol = 1,
    scales = "free_y"
  ) +
  scale_color_manual(
    values = c(
      better = "#2C6E9B",
      worse = "#C47B22"
    ),
    labels = c(
      better = "Better sleep",
      worse = "Worse sleep"
    )
  ) +
  scale_x_date(
    breaks = breaks_pretty(n = 6),
    date_labels = "%b %d"
  ) +
  labs(
    title = paste(
      "Sleep and next-day glucose for participant",
      example_participant
    ),
    subtitle = str_wrap(
      "Participant selected by the greatest number of eligible nights; dashed line is their median sleep duration.",
      width = 90
    ),
    x = "Wake-up date",
    y = NULL,
    color = NULL
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

# Results in plain language

```{r}
median_glucose_difference <- median(
  paired_differences$mean_glucose_difference,
  na.rm = TRUE
)

glucose_difference_iqr <- quantile(
  paired_differences$mean_glucose_difference,
  c(0.25, 0.75),
  na.rm = TRUE
)

share_positive_difference <- mean(
  paired_differences$mean_glucose_difference > 0,
  na.rm = TRUE
)

median_sleep_gap <- median(
  paired_differences$sleep_duration_gap_hours,
  na.rm = TRUE
)
```

The paired analysis retained `r nrow(paired_differences)` participants and `r nrow(analysis_nights)` person-nights. On average within a participant's two categories, better-sleep nights were separated from worse-sleep nights by a median of `r round(median_sleep_gap, 1)` hours.

The median within-person next-day mean-glucose difference was `r round(median_glucose_difference, 1)` mg/dL (interquartile range `r round(glucose_difference_iqr[1], 1)` to `r round(glucose_difference_iqr[2], 1)` mg/dL), and `r percent(share_positive_difference, accuracy = 1)` of paired participants had a positive difference. These values describe this small public-data sample; they are not an estimate of a causal sleep effect.

# Limitations

This analysis is observational and supports **no causal claim**: shorter sleep may accompany stress, illness, schedule changes, or other factors that may also affect glucose. The public data does not provide meal timestamps, so post-meal glucose responses cannot be separated from other next-day variation. Medication timing is not fully available in the public data, which prevents alignment of doses with the CGM window. Dexcom CGM and Garmin sleep/activity measurements are consumer or wearable measurements rather than laboratory gold standards; Garmin sleep stages in particular are imperfect estimates. Finally, AI-READI is cross-sectional, with only a short monitoring period per participant, so the analysis cannot establish long-term direction, prediction, or clinical outcomes.
