---
title: "R Documentation Notes"
author: "Tyelr Tong"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
---
# Basic Info
## Datatypes
 - Integers: Integer values
 - Doubles/Numerics: Decimal values
 - Logicals: Booleans, TRUE or FALSE
 - Characters: A set of characters
 - Vectors: a 1D set of values. Created using the c() function.
 - Factors: Categorical data, represented as strings.
 - Data Frames: A 2D set of data in the shape of a rectangle. Created with the data.frame() function

```{r}
i = 1
# or
i <- 1 
# i is a integer with value 1
d <- 3.14
# d is a double with value 3.14
l <- TRUE
# l is a logical with value TRUE
c <- "hello world"
# c is a character with value "hello world"
v <- c(1, 2, 3, 4, 5)
# v is a vector containing the set 1, 2, 3, 4, 5
d <- data.frame(
  Integers = c(1, 2, 3),
  Chars = c("A", "B", "C"),
  Logicals = c(TRUE, FALSE, TRUE)
)
# d is a 3x3 data frame with three variables, each with three values.
```

## Conditionals
 - & for AND and | for OR
 - ==, <=, >=, <, < for comparisons, they all return Logicals
```{r}
a <- FALSE
b <- TRUE
c <- a & b
# c is FALSE
c <- a | b
# c is TRUE
i <- 5
j <- 10
k <- i > j
# k is FALSE
#l <- i = j
#This throws an error because = is used for assigning values to things. Instead:
l <- i == j
# l is FALSE
```

## Functions
 - Do things in R.

```{r}
```

 - Some have default values that can be changed.

```{r}
```
 
## Errors, Warnings, and Messages
 - All errors, warnings and messages are in red.
 - Errors all start with "Error in ..." followed by where the error is and a explanation for why. Code will NOT work when there are errors.
 - Warnings all start with "Warning: " and then explain the warning. In general, code will still work, but there will be caveats, generally explained in the warning.
 - If there is red text that does not start with "Error in ..." or "Warning: ", it is a message. Messages are in general harmless diagnostics.

## Packages
 - Extend functionality by providing new functions, data, and documentation.
 - Packages can be installed using the Packages tab in R Studio.
 - Packages can also be installed with "install.packages("package_name_here")".
 - Packages must be loaded after installing to use them.
 - This can be done with "library("package_name_here")"
 - You MUST load the package you want to use EVERY TIME you open RStudio.

```{r, message=FALSE}
# install.packages("tidyverse")
library(tidyverse)
```
 
## More on Data Frames
 - The View() function, which is built in to R, opens RStudio's data viewer when called on a data frame.
 - The glimpse() function, provided by the dplyr package, prints a representation of the data frame provided. 
 - The kable() function, provided by the knitr package, also prints a representation of the data frame provided, but the output is more reader-friendly than that of the glimpse() function.
 - The $ operator lets you get all the values of a variable from a data frame.
 
# The Tidyverse
## Data Visualization
### The Grammar of Graphics
 - The grammar of graphics is a theoretical framework for data visualization, serving as the foundation of the package ggplot2.
 - In languages, there are multiple components to a sentence. These components must be put together in a certain way in order to construct a correct sentence. The grammar of graphics is like a set of rules, just like how grammar functions in language, to put pieces of a infographic together.
 
### The Components of Grammar
 - data: the dataset containing the variable of interest.
 - geom: the geometric object in question
 - aes: the aesthetic attributes of the geometric object. Includes position, color, shape, and size. Aesthetic attributes are mapped to variables in the dataset.

### The Five Named Graphs
